\documentclass[preprint,1p,a4paper,11pt]{elsarticle}

\input{preamble.tex}
\geometry{margin=25mm,
    headheight=110pt,
    footskip=30pt
}
\biboptions{sort&compress}

% We need slightly different text in the manual and in the release note.
\newif\ifreleasenote
\releasenotetrue % comment out to hide answers

\journal{European Physical Journal C}

\begin{document}
\begin{frontmatter}
\begin{flushright}
CERN-TH-2023-237,  MPP-2023-285, OUTP-23-15P
\end{flushright}
\title{\hoppet{} {\tt v2.0.0} release note}

\author[1]{Alexander Karlberg}\ead{alexander.karlberg@cern.ch}
\author[2]{Paolo Nason}\ead{paolo.nason@mib.infn.it}
\author[3,4]{Gavin Salam}\ead{gavin.salam@physics.ox.ac.uk}
\author[5,6]{Giulia Zanderighi}\ead{zanderi@mpp.mpg.de}
\author[3]{Fr\'ed\'eric Dreyer\gps{what address?}}\ead{frederic.dreyer@gmail.com}

\affiliation[1]{organization={CERN, Theoretical Physics Department}, postcode={CH-1211} ,city={Geneva 23}, country={Switzerland}}
\affiliation[2]{organization={INFN, Sezione di Milano-Bicocca, and Universita di Milano-Bicocca, \mbox{Piazza della Scienza 3}}, postcode={20126} ,city={Milano}, country={Italy}}
\affiliation[3]{organization={Rudolf Peierls Centre for Theoretical Physics, Clarendon Laboratory, Parks Road}, postcode={OX1 3PU} ,city={Oxford}, country={UK}}
\affiliation[4]{organization={All Souls College}, postcode={OX1 4AL} ,city={Oxford}, country={UK}}
\affiliation[5]{organization={Max-Planck-Institut fur Physik, Boltzmannstr. 8}, postcode={85748} ,city={Garching}, country={Germany}}
\affiliation[6]{organization={Physik-Department, Technische Universitat Munchen, James-Franck-Strasse 1}, postcode={85748} ,city={Garching}, country={Germany}}

\begin{abstract}
  We document the three main new features in release 2.0.0 of the
  \hoppet parton distribution function evolution code, specifically
  support for QED evolution to an accuracy phenomenologically
  equivalent to NNLO QCD, for N$^3$LO QCD evolution in the variable
  flavour number scheme, and for the determination of hadronic
  structure functions for massless quarks up to N$^3$LO. Additionally
  we describe a new Python interface, CMake build option, and
  functionality to save a \hoppet{} table as an LHAPDF grid.
\end{abstract}

\begin{keyword}
  Perturbative QCD \sep DIS \sep DGLAP \sep QED
%% keywords here, in the form: keyword \sep keyword

%% PACS codes here, in the form: \PACS code \sep code

%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)

\end{keyword}

\end{frontmatter}
\tableofcontents
\section{Introduction}


\hoppet~\cite{Salam:2008qg} is a parton distribution function (PDF)
evolution code written in modern Fortran,\footnote{This release of
\hoppet makes use of features introduced in Fortran~2008, for example
its \ttt{abstract interface}, and hence a Fortran~2008 compliant
compiler is now needed. The code has been tested to compile and run
with \texttt{gfortran v9.5.0} and later and the 2025 version of the
Intel compiler \texttt{ifx}.}  with interfaces also for C/\CPP{} and
earlier dialects of Fortran.
%
It offers both a high-level PDF evolution interface and user-access to
lower-level functionality for operations such as convolutions of
coefficient functions and PDFs.
%
It is designed to provide flexible, fast and accurate evolution.

Since the first major release of \hoppet, the landscape of PDF
evolution codes has evolved substantially, with a range of new open-source
codes having been developed, for example the APFEL~\cite{Bertone:2013vaa}, \APFELPP{}~\cite{Bertone:2017gds} and EKO~\cite{Candido:2022tld}
codes (the latter including
features such as cached evolution that have long been standard in
\hoppet).\footnote{Other codes such as ChiliPDF~\cite{Diehl:2021gvs}
  appear to not yet be public.}
%
Nevertheless, \hoppet remains a powerful tool, notably for its ability
to reach high and quantifiable accuracies with competitive speed.
%
As a result it often provides a critical reference for benchmark
studies, as those presented e.g.\ in
Refs.~\cite{Dittmar:2005ed,Bertone:2024dpm}.
%
It also provides the option of fast (millisecond-scale) evolution with
accuracies $\sim 10^{-4}$, which is more than sufficient for
phenomenological applications.
%
%
Furthermore its exposition not just of full PDF evolution but also of
low-level functionality makes it useful for a range of fixed-order,
all-order and Monte Carlo studies\ak{Worth making references to use in panscales, powheg, various resummation tools?}.
%
With the advent of ever more precise data from the LHC at CERN, and
future very high precision data at the forthcoming Electron-Ion
Collider~\cite{AbdulKhalek:2021gbh} at Brookhaven National Laboratory,
the continued development of tools such as \hoppet remains important
for the field.

This release note documents three major additions to \hoppet, made
available as part of release~2.0.0: (1) support for QED evolution,
originally developed as part of the LuxQED project for the evaluation
of the photon density inside a proton and its extension to lepton
distributions in the
proton~\cite{Manohar:2016nzj,Manohar:2017eqh,Buonocore:2020nai,Buonocore:2021bsf};
%
(2) support for QCD evolution at N$^3$LO in the (zero mass) variable
flavour number scheme (VFNS)~\cite{Buza:1996wv};
%
and (3) support for the
determination of massless hadronic structure functions, as initially developed
for calculations of vector-boson fusion cross
sections~\cite{Cacciari:2015jma,Dreyer:2016oyx,Dreyer:2018qbw,Dreyer:2018rfu}.

This release also includes a range of other smaller additions relative
to the original 1.1.0 release documented in \cite{Salam:2008qg}, in
particular the addition of a new Python interface and CMake build
option.
%
Unified documentation of the whole \hoppet package is part of the
distribution at \url{https://github.com/hoppet-code/hoppet} in the
\masterlink{doc/} directory.
%
Details of the other changes since release 1.1.0 can be found in the
\masterlink{NEWS} and \masterlink{ChangeLog} files from the
repository.\medskip

\noindent \fbox{\parbox{\textwidth}{
\textbf{Note that for this 2.0.0 release, \hoppet's library name has been renamed
from \texttt{hoppet\_v1} to \texttt{hoppet}, and similarly for the main
module and \CPP include file.
%
Users with an existing installation of \hoppet (v1) should make sure
that they link with the new library name.}
}}

\input{qed-and-structure-functions.tex}

%======================================================================
\section{Conclusion}

The \hoppet additions described here provide fast access to QED
evolution and to structure-function calculations, including evolution
in the VFNS at N$^3$LO, which we hope may be of benefit beyond the
projects where they have already been used.


\section*{Acknowledgments}

We gratefully acknowledge the work of Luca Buonocore
for the implementation of the $P_{lq}$ splitting function in the QED
code,
%
and collaboration with Melissa van Beekveld on options for speed
improvements in the evaluation of tabulated PDFs.
%
We are grateful to Johannes Bl\"umlein for providing us with a
pre-release version of the code in Ref.~\cite{BlumleinCode} and a
suitable license for its use.
%
We also wish to thank Juan Rojo for useful discussions. 
%
GPS wishes to acknowledge funding from a Royal Society Research
Professorship (grant RP$\backslash$R$\backslash$231001) and from the Science and
Technology Facilities Council (STFC) under grant ST/X000761/1.
%
PN thanks the Humboldt Foundation for support. 
%
\appendix

\input{perturbative-evolution-in-qcd.tex}

\bibliographystyle{elsarticle-num}
\bibliography{hoppet.bib}
%\input{bibliography.tex}

\end{document}

